COFFEE: an Optimizing Compiler for Finite Element Local Assembly

### 2014-06-29 TACO-paper/coffee.tex
edits

% At a minimum you need to supply the author names, year and a title.
% IMPORTANT:
% Full first names whenever they are known, surname last, followed by a period.
% In the case of two authors, 'and' is placed between them.
% In the case of three or more authors, the serial comma is used, that is, all author names
% except the last one but including the penultimate author's name are followed by a comma,
% and then 'and' is placed before the final author's name.
% If only first and middle initials are known, then each initial
% is followed by a period and they are separated by a space.
% The remaining information (journal title, volume, article number, date, etc.) is 'auto-generated'.

@@

\section{Introduction}
In many fields, such as computational fluid dynamics, computational
electromagnetics, and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely
employed to approximate solutions of these PDEs. Unstructured meshes
are often used to discretize the equation domain, since their
geometric flexibility allows solvers to be extremely effective. The
solution is sought by applying suitable numerical operations, or
kernels, to the components of a mesh, such as edges, vertices, or
cells. On standard clusters of multi-cores, typically, a kernel is
executed sequentially by a thread, while parallelism is achieved
partitioning the mesh and assigning each partition to a different node
or thread. Such an execution model, with minor variations, is adopted,
for example, in \cite{pyop2isc}, \cite{Fenics},
\cite{fluidity_manual_v4}, \cite{lizst}.

@@

Our experiments show that the original generated code for non-trivial
assembly kernels, despite following state-of-the-art techniques,
remains suboptimal in the context of modern multi-core
architectures. Our cost-model-driven sequence of code transformations,
aimed at improving SIMD vectorization and register data locality, can
result in performance improvements up to 4.4$\times$ over original
kernels, and around 70$\%$ of the test cases obtain a speed up greater
than 2$\times$. The contributions of this paper, therefore, are
\begin{enumerate}
\item An optimization strategy for finite element local assembly. Our
  approach exploits domain knowledge and goes beyond the limits of
  both vendor and research compilers.
\item Design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item Systematic analysis using a suite of examples of real-world
  importance, and evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{enumerate}

>>>>

% At a minimum you need to supply the author names, year and a title.
% IMPORTANT: Full first names whenever they are known, surname last,
% followed by a period.  In the case of two authors, 'and' is placed
% between them.  In the case of three or more authors, the serial
% comma is used, that is, all author names except the last one but
% including the penultimate author's name are followed by a comma, and
% then 'and' is placed before the final author's name.  If only first
% and middle initials are known, then each initial is followed by a
% period and they are separated by a space.  The remaining information
% (journal title, volume, article number, date, etc.) is
% 'auto-generated'.

@@

\section{Introduction}
In many fields such as computational fluid dynamics, computational
electromagnetics and structural mechanics, phenomena are modelled by
partial differential equations (PDEs). Numerical techniques, like the
finite volume method and the finite element method, are widely
employed to approximate solutions of these PDEs. Unstructured meshes
are often used to discretize the equation domain, since their
geometric flexibility allows solvers to be extremely effective. The
solution is sought by applying suitable numerical operations, or
kernels, to the components of a mesh, such as edges, vertices, or
cells. On standard clusters of multicores, typically, a kernel is
executed sequentially by a thread, while parallelism is achieved
partitioning the mesh and assigning each partition to a different node
or thread. Such an execution model, with minor variations, is adopted,
for example, in \cite{pyop2isc}, \cite{Fenics},
\cite{fluidity_manual_v4}, \cite{lizst}.

@@

Our experiments show that the original generated code for non-trivial
assembly kernels, despite following state-of-the-art techniques,
remains suboptimal in the context of modern multicore
architectures. Our domian-aware cost-model-driven sequence of code
transformations, aimed at improving SIMD vectorization and register
data locality, can result in performance improvements up to
4.4$\times$ over original kernels, and around 70$\%$ of the test cases
obtain a speed up greater than 2$\times$. The contributions of this
paper, therefore, are
\begin{enumerate}
\item An optimization strategy for finite element local assembly. Our
  approach exploits domain knowledge and goes beyond the limits of
  both vendor and research compilers.
\item Design and implementation of a compiler that automates the
  proposed code transformations for any problems expressible in
  Firedrake.
\item Systematic analysis using a suite of examples of real-world
  importance, and evidence of significant performance improvements on
  two Intel architectures, a Sandy Bridge CPU and the Xeon Phi.
\end{enumerate}
